Analyzing Fairness: Race, Recidivism, and the COMPAS Algorithm

### 2022-04-25 final_metrics_analysis.Rmd

# for Amrita
use_condaenv("anaconda3")
```

@@

# from aif360.algorithms.preprocessing import DisparateImpactRemover
# from BlackBoxAuditing.repairers.GeneralRepairer import Repairer

@@

# Exploratory Data Analysis
```{python EDA} 
# needs interpretation
X_new_index = X_new.rename(columns={"race": "def_race"})
X_new_index = X_new_index.rename(columns={"sex": "def_sex"})
# renaming dataframe columns to avoid the ValueError of variables being a column AND index label (which is ambiguous)

@@

@@

1242 (59.14%) of white defendants were charged with felonies versus 2194 (69.15%) of Black defendants. Additionally, a higher percentage of white defendants (40.86%) were charged with a misdemeanor than black defendants (30.85%).

@@

```{python exploratory age by race plot}
# age distribution by race - hacking solution for purple dotted line for now
#ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
#kdeline = ax.lines[0]
#mean_black = df_viz.groupby('race').age.median()[0]
#xs = kdeline.get_xdata()
#ys = kdeline.get_ydata()
#height_black = 0.027 # np.interp(mean_black, xs, ys)
#ax.vlines(mean_black, 0, height_black, color=purple, ls=':')

#mean_white = df_viz.groupby('race').age.median()[1]
#height_white = np.interp(mean_white, xs, ys)
#ax.vlines(mean_white, 0, height_white, color=green, ls=':')

#plt.title('Distribution of Age by Race')
```

@@

@@

# Pre-Processing Approaches

@@

```{python pre-processing disparate impact remover}
#lr = LogisticRegressionCV(solver='lbfgs')
#rew = ReweighingMeta(estimator=lr, reweigher=Reweighing('race'))
#rew.fit(X_train, y_train)
#y_pred_REW = rew.predict(X_test)

# diremover = DisparateImpactRemover(repair_level=1.0, sensitive_attribute="race")
#diremover.fit(X_train, y_train)
#y_pred_diremover = diremover.predict(X_test)

# diremover_rp = diremover.fit_transform(X_test)
# AttributeError: 'DataFrame' object has no attribute 'features' (probem with .fit_transform function)
# .fit_transform() angry because of X_test... 
```

# In-Processing Approaches

@@

## Adversarial Debiasing
```{python adversarial debiasing}
# from aif360.algorithms.inprocessing import AdversarialDebiasing
# import tensorflow as tf
# 
# adv_deb = AdversarialDebiasing(unpr)
# #adv_deb.fit(X_train, y_train)
# #y_pred_AD = adv_deb.predict(X_test)
# #adv_deb.sess_.close()
```

@@

The highest percentage in this confusion matrix is for false positives, defendants who the model predicted to recidivate and actually did not, at 37.76%. The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is lower at 23.88%. The percentage of false negatives is lower, whereas the true negative is low at 14.33%. This indicates that the model does not do a good job in predicting the defendents who don't recidivate. It is very likely that it will incorrectly predict that someone will recidivate.

@@

@@

Compared to the baseline model, the predjudice remover is highly in favor of the unprivileged group because its value is greater than 0. The value of .367 is much higher than 0, the value of fairness.

@@

Compared to the baseline model, the predjudice remover is highly in favor of the unprivileged group because its value is greater than 0. The value of .321 is much higher than 0, the value of fairness.

@@

Compared to the baseline model, the predjudice remover is highly in favor of the unprivileged group because its value is greater than 0.

@@

@@

The highest percentage in this confusion matrix is for true negatives, defendants who the model predicted to not recidivate and actually did not, at 50.04%. The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is much lower at 3.11%. The percentage of false negatives is somewhat high for this baseline model, whereas the false positive rate is very low at 2.05%. This indicates that the model does a good job in predicting the defendents who don't recidivate. It is very unlikely that it will incorrectly predict that someone will recidivate.

```{python}
#metrics_per_group(y_test, y_pred_CEO)
#plt.show()
```

```{python}
def get_attributes(data, selected_attr=None):
    unprivileged_groups = []
    privileged_groups = []
    if selected_attr == None:
        selected_attr = data.protected_attribute_names

    for attr in selected_attr:
            idx = data.protected_attribute_names.index(attr)
            privileged_groups.append({attr:data.privileged_protected_attributes[idx]}) 
            unprivileged_groups.append({attr:data.unprivileged_protected_attributes[idx]})

    return privileged_groups, unprivileged_groups
```

```{python post-processing equal odds needs fix} 
#privileged_groups, unprivileged_groups = get_attributes(train_BLD, selected_attr= 'race')
#ep = EqOddsPostprocessing(unprivileged_groups= unprivileged_groups, privileged_groups= privileged_groups, seed = 42)
#eo = PostProcessingMeta(estimator=lr, postprocessor=ep, random_state=1234567)
#EOPP = ep.fit(train_BLD, test_BLD)
#y_pred_EO = ep.predict(test_BLD)
#y_proba_EO = ep.predict_(test_BLD)

#acc_EO = accuracy_score(y_test, y_pred_EO)
```

@@

At the value of 0.134, the statistical parity difference is now positive and slightly in favor of black defendants.

@@

There has been a significant change from -.408 to 0.088, showing that now the value is within the bounds of "fair" (-0.1 to 0.1)

@@

The baseline model with the average odds difference was -0.438 and now it is 0.15 which indicates being slightly in favor of black defendants.

@@

The baseline value was 0.466 which is highly out of the range of "fair" which goes up to 1. Now at a value of 1.2 it is much higher than 1.

@@

>>>>

# for Amrita
#use_condaenv("anaconda3")
```

@@

@@

# Exploratory Data Analysis

### Descriptions of Average Defendants
```{python EDA} 
# needs interpretation
X_new_index = X_new.rename(columns={"race": "def_race"})
X_new_index = X_new_index.rename(columns={"sex": "def_sex"})
# renaming dataframe columns to avoid the ValueError of variables being a column AND index label (which is ambiguous)

@@

### Recidivism Distribution

### Racial Distributions

#### Age

```{python exploratory age by race plot}
# age distribution by race - hacking solution for purple dotted line for now
ax = sns.kdeplot(x="age", hue="race", data=df_viz, palette=workshop_palette);
kdeline = ax.lines[0]
mean_black = df_viz.groupby('race').age.median()[0]
xs = kdeline.get_xdata()
ys = kdeline.get_ydata()
height_black = 0.027 # np.interp(mean_black, xs, ys)
ax.vlines(mean_black, 0, height_black, color=purple, ls=':')

mean_white = df_viz.groupby('race').age.median()[1]
height_white = np.interp(mean_white, xs, ys)
ax.vlines(mean_white, 0, height_white, color=green, ls=':')

plt.title('Distribution of Age by Race')
```

#### Charge Degree

@@

1242 (59.14%) of white defendants were charged with felonies versus 2194 (69.15%) of black defendants. Likewise, a higher percentage of white defendants (40.86%) were charged with a misdemeanor than black defendants (30.85%).

#### Charge Descriptions

@@

#### Recidivism

@@

#### Prior Charges

@@

As discussed above, all four fairness metrics define the baseline model as unfairly benefitting white defendants. The green bar represents the range of fair values, while the blue bars are the actual metric values for the baseline model. We can see that the margin of unfairness is smallest for statistical parity difference, as the value is only 0.038 outside of the -0.1 to 0.1 range of fairness. However, for the other three metrics, the value is significantly below the margin of fairness.

# Pre-Processing

@@

The orange bars in these graphs represent the new values of the fairness metrics after implementing the reweighting approach. Like we saw above, reweighting the data improved the fairness in all four of these metrics. All four values are now within the range of values considered fair. Statistical parity difference and disparate impact ratio still imply a slight benefit for white defendants, while equal opportunity difference and average odds difference now imply a slight benefit for Black defendants.

# In-Processing

@@

@@

The highest percentage in this confusion matrix is for false positives, defendants who the model predicted to recidivate and actually did not, at 37.76%. This false positive percentage concerns us because we do not want defendants who do not recidivate to have unfairly long sentences due to their (incorrectly) predicted recidivism. The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is lower at 23.88%. The percentage of false negatives is lower, whereas the true negative is low at 14.33%. This indicates that the model does not do a good job in predicting the defendents who don't recidivate. It is very likely that it will incorrectly predict that someone will recidivate.

@@

After the prejudice remover, the accuracy of the model is fairly similar across the board of races, but the model's accuracy is only 38.21%, which is much lower than the baseline and reweighting models. This accuracy score means that the model makes accurate predictions only 38% of the time.

In terms of recall, the score for white defendants is very high at 0.7, meaning that the model correctly identifies white defendants who recidivate about 70% of the time. This number is about 50% for all defendants, but only 40% for Black defendants. Just as we saw a low recall score for white defendants in the baseline model, now we see the same for Black defendants in the prejudice remover model. This means that the model frequently predicts Black defendants to not recidivate (the more favorable outcome) when they actually do.

Finally, the precision score has lowered from about 0.6 for the baseline model to about 0.4 for all defendants, 0.45 for Black defendants, and 0.3 for white defendants. This means that the model's predictions of a defendant who recidivates are accurate about 40% of the time, a number which is slightly higher for Black defendants and slightly lower for white defendants.

@@

Compared to the baseline model, the predjudice remover is highly in favor of the unprivileged group, Black defendants, because its value is greater than 0. The value of 0.367 is much higher than 0, the value of fairness.

@@

Compared to the baseline model, the predjudice remover is highly in favor of Black defendants because its value is greater than 0. The value of 0.321 is much higher than 0, the value of fairness.

@@

Compared to the baseline model, the predjudice remover is highly in favor of Black defendants, because its value is greater than 0.

@@

Like the model performance metrics suggested, the prejudice remover approach resulted in an increased benefit for Black defendants. As the orange bars on these plots show, the values of the fairness metrics have reversed from their baseline values. Now all four metrics suggest an unfair advantage for Black defendants. Thus, this approach removed the model's prejudice against Black people, but it did not result in a "fair" model.

@@

The highest percentage in this confusion matrix is for true negatives, defendants who the model predicted to not recidivate and actually did not, at 50.04%. The number of true positives, defendants who the model predicted to recidivate and actually recidivated, is much lower at 3.11%. The percentage of false negatives is somewhat high for this baseline model, whereas the false positive rate is very low at 2.05%. This indicates that the model does a good job in predicting the defendents who don't recidivate. It is very unlikely that it will incorrectly predict that someone will recidivate. Though the accuracy score for the model is 0.53, lower than the baseline, this model is much better at identifying true negatives.

@@

At the value of 0.134, the statistical parity difference is now positive and slightly in favor of Black defendants. However, it is just outside the bounds of fairness.

@@

There has been a significant change from -0.408 to 0.088, showing that now the value is within the bounds of "fair" (-0.1 to 0.1). Additionally, since the value is positive, it implies a slight benefit for Black defendants.

@@

The baseline model with the average odds difference was -0.438 and now it is 0.15, which indicates slight unfairness (as it is above 0.1) and the model being in favor of Black defendants.

@@

The baseline value was 0.466, which is significantly below the range of "fair" which begins at 0.8. Now at a value of 1.15 it is above 1, conveying fairness and a benefit of Black defendants.

@@

As we can see from the orange bars, the calibrated equalized odds approach results in all four metrics now suggesting a benefit for the originally unprivileged group, Black defendants. Though the values for statistical parity difference and average odds difference are slightly above the range of what is considered fair, the margin is much smaller than the original margin between the value and the range of fairness (as indicated by the blue bars). Thus, the calibrated equalized odds approach successfully counteracts the bias against Black defendants and results in a mostly fair model.
